Algorithms for Longest Common Extensions

### 2011-07-18 doc/report.tex
m

Given a string $s$, the fingerprint $F_t[i]$ is a natural number identifying the substring $s[i\twodots i+t]$ among all $t$-length substrings of $s$. We assign fingerprints such that for any $i$, $j$ and $t$, $F_t[i] = F_t[j]$ if and only if $s[i\twodots i+t] = s[j\twodots j+t]$. In other words, if two substrings of $s$ have the same length, they have the same fingerprints if and only if the substrings themselves are the same.

At the end of a string when $i+t>n$, we define $F_t[i]$ by adding extra characters to the end of the string as needed.

@@

To perform an LCE query using the \fprintk\ data structure, start with $v=0$ and $\ell=0$, then do the following steps:
\begin{samepage}
\begin{enumerate}
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless and $\ell = k-1$.
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$ and repeat this step.
\item When $H_\ell[i+v] \neq H_\ell[j+v]$, decrement $\ell$ by one and repeat step 2.
\item Stop when $\ell = 0$ and $H_\ell[i+v] \neq H_\ell[j+v]$.
\end{enumerate}

@@

We can preprocess the data structure in $O(\sortt + k\cdot n)$ time using the SA and LCP arrays. First create the SA and LCP arrays. Then preprocess each of the $k$ levels using the following steps:
\begin{samepage}
\begin{enumerate}
\item Loop through the $n$ substrings of length $t_\ell$ in lexicographically sorted order by looping through the elements of SA.
\item Assign an arbitrary fingerprint to the first substring.
\item If the current substring $s[\SA[i]\twodots\SA[i]+t_\ell]$ is equal to the substring examined in the previous iteration of the loop, give the current substring the same fingerprint as the previous substring, otherwise give the current substring a new unused fingerprint. The two substrings are equal when $\LCE[i] \geq t_\ell$.
\end{enumerate}

@@

\begin{theorem}
\label{thm:lz-tree}
Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$. We can preprocess $z$ using $O(n)$ space to answer LCE queries $v=\LCE(i,j)$ in $O(\log\log N + \log n \cdot \textit{\#phrases})$ time, where \textit{\#phrases} is the number of phrases of $z$ used to store $s[i\twodots i+v]$ and $s[j\twodots j+v]$.
\end{theorem}

@@

\newcommand{\zlength}{\textit{length}}
\newcommand{\zindex}{\textit{index}}
\newcommand{\zend}{\textit{end}}

To achieve the result of \pref{thm:lz-tree}, we need to do LCE on a tree in $O(\log n)$ query time and $O(n)$ space. In this section we analyze ways to achieve this.

@@

\paragraph{Paths}
A path $p=(u,w)$ in a tree is a path, which starts at node $u$ and ends at node $w$, where $u$ is an ancestor of $w$. We use the following notation on path $p=(u,w)$:
\begin{itemize}
\item The path has length $e=\zlength(p)=d(w)-d(u)$, i.e.\ $p$ has $e$ edges and $e+1$ nodes.
\item The node $p[a]$ is a node in $p$, such that the distance between $u$ and $p[a]$ is $a$. The start and end nodes of the path are $p[0]=u$ and $p[e]=w$. We can calculate $p[a]$ in constant time as $p[a]=\LA(d(u)+a,w)$, where $d(u)$ is the depth of $u$, and $\LA$ is a Level Ancestor query.
\item The subpath $p[a\twodots b]=(p[a],p[b])$ is the path, which starts at node $p[a]$ and ends at node $p[b]$. The entire path is $p = p[0,e]$.
\item The label of a path $\zlabel(p)=\zlabel(p[1])\cdot...\cdot\zlabel(p[e])$ is the concatenation of the labels on all edges between nodes in the path.
\end{itemize}

\paragraph{LCE on trees}
Given two paths $p_i=(u_i,w_i)$ and $p_j=(u_j,w_j)$ of lengths $e_i$ and $e_j$ in a tree $T$, the Longest Common Extension $v=\LCE_T(p_i, p_j)$ is the length of the longest common prefix of $\zlabel(p_i)$ and $\zlabel(p_j)$.

@@

\newcommand{\tfprint}[1][k]{\ensuremath{\proc{TreeFingerprint}_{#1}}}
\newcommand{\tfprintk}{\tfprint[k]}

@@

@@

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.7\textwidth,page=1]{tree-fingerprint.pdf}
    \end{center}
    \caption{\label{fig:tree-query}Finding the right fingerprint to look at for a \tfprintk\ query.}
\end{figure}

The fingerprints we choose to compare are the fingerprints, which represents substrings on $p_i$ and $p_j$. At any given iteration at level $\ell$, we know that $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and we now check if $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$, as shown in \pref{fig:tree-query}. From our definition of fingerprints in a tree, these two labels have fingerprints $H_\ell[p_i[v+t_\ell]]$ and  $H_\ell[p_j[v+t_\ell]]$, which is what our LCE query uses.
\end{proof}

@@

In a \proc{HeavyPathLce} query $\LCE_T(p_i,p_j)$ we start with $v=0$ and do the following repeatedly:
\begin{enumerate}
\item Stop when we reach either $v=e_i$ or $v=e_j$, and return $v$.
\item Find $m_i = \zlength(d_i[v], \NCA_T(\zend(p_i[v+1]), w_i))$ and $i=\zindex(p_i[v+1])$.
\item Find $m_j = \zlength(d_j[v], \NCA_T(\zend(p_j[v+1]), w_j))$ and $j=\zindex(p_j[v+1])$.
\item Find $v'=\min\{\LCE_s(i,j),m_i,m_j\}$.
\item Increment $v$ by $v'$.
\item Stop if $v'<\min\{m_i,m_j\}$ and return $v$.
\end{enumerate}

@@

The \proc{LadderLce} query starts with $v=0$ and repeats the following:
\begin{enumerate}
\item Stop when $v = e_i$ or $v = e_j$ and return $v$.
\item Find $m_i = \zlength(p_i[v\twodots e_i])/2$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
\item Find $m_j = \zlength(p_j[v\twodots e_j])/2$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.
\item Find $v' = \min\{LCE_s(i,j),m_i,m_j\}$
\item Increment $v$ by $v'$.
\item Stop if $v' < \min\{m_i,m_j\}$ and return $v$.
\end{enumerate}

>>>>

Given a string $s$, the fingerprint $F_t[i]$ is a natural number identifying the substring $s[i\twodots i+t-1]$ among all $t$-length substrings of $s$. We assign fingerprints such that for any $i$, $j$ and $t$, $F_t[i] = F_t[j]$ if and only if $s[i\twodots i+t-1] = s[j\twodots j+t-1]$. In other words, if two substrings of $s$ have the same length, they have the same fingerprints if and only if the substrings themselves are the same.

At the end of a string when $i+t-1>n$, we define $F_t[i]$ by adding extra characters to the end of the string as needed.

@@

To perform a LCE query using the \fprintk\ data structure, start with $v=0$ and $\ell=0$, then do the following steps:
\begin{samepage}
\begin{enumerate}
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless and $\ell = k-1$.
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$ and repeat this step.
\item When $H_\ell[i+v] \neq H_\ell[j+v]$, decrement $\ell$ by one and repeat step 2.
\item Stop when $\ell = 0$ and $H_\ell[i+v] \neq H_\ell[j+v]$.
\end{enumerate}

@@

We can preprocess the data structure in $O(\sortt + k\cdot n)$ time using the SA and LCP arrays. First create the SA and LCP arrays. Then preprocess each of the $k$ levels using the following steps:
\begin{samepage}
\begin{enumerate}
\item Loop through the $n$ substrings of length $t_\ell$ in lexicographically sorted order by looping through the elements of SA.
\item Assign an arbitrary fingerprint to the first substring.
\item If the current substring $s[\SA[i]\twodots\SA[i]+t_\ell-1]$ is equal to the substring examined in the previous iteration of the loop, give the current substring the same fingerprint as the previous substring, otherwise give the current substring a new unused fingerprint. The two substrings are equal when $\LCE[i] \geq t_\ell$.
\end{enumerate}

@@

\begin{theorem}
\label{thm:lz-tree}
Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$. We can preprocess $z$ using $O(n)$ space to answer LCE queries $v=\LCE(i,j)$ in $O(\log\log N + \log n \cdot \textit{\#phrases})$ time, where \textit{\#phrases} is the number of phrases of $z$ used to store $s[i\twodots i+v-1]$ and $s[j\twodots j+v-1]$.
\end{theorem}

@@

\newcommand{\zlength}{\textit{length}}
\newcommand{\tfprint}[1][k]{\ensuremath{\proc{TreeFingerprint}_{#1}}}
\newcommand{\tfprintk}{\tfprint[k]}
\newcommand{\zindex}{\textit{index}}
\newcommand{\zend}{\textit{end}}

To achieve the result of \pref{thm:lz-tree}, we need to do LCE on a tree in $O(\log n)$ query time and $O(n)$ space. In this section we analyze ways to achieve this. We first define LCE on trees, and then describe different solutions we have found to the problem. \pref{tab:tree-lce-algs} lists the different solutions we have found.

\begin{table}[tp]
\centering
\begin{tabular}{l|l|l|l}
\hline\hline
Algorithm & Space & Query time & Preprocessing \\ [0.5ex] \hline
\proc{TreeDirectComp} & $O(n)$ & $O(n)$ & $O(n)$ \\ \hline
\tfprintk, $1\leq k\leq\logceil$ & $O(kn)$ & $O(k\cdot n^{1/k})$ & $O(n\log n)$ \\ \hline
\proc{HeavyPathLce} & $O(n)$ & $O(\log n)$ & $O(\sortt)$ \\ \hline
\proc{TreeLcpRmq} & $O(n^2)$ & $O(1)$ & - \\ \hline
\proc{TreeSuffixNca} & $O(n^2)$ & $O(1)$ & - \\ \hline
\proc{LadderLce} & $O(n)$ & $O(\log n)$ & $O(\sortt)$ \\ \hline
\end{tabular}
\caption{Algorithms for LCE on trees with their space requirements, query times and preprocessing times.}\label{tab:tree-lce-algs}
\end{table}

@@

\paragraph{Paths}
A path $p=(u,w)$ in a tree is a path, which starts at node $u$ and ends at node $w$, where $u$ is an ancestor of $w$. We use the following notation on paths:
\begin{itemize}
\item The path $p=(u,w)$ has length $e=\zlength(p)=d(w)-d(u)$, i.e.\ $p$ has $e$ edges and $e+1$ nodes.
\item The node $p[x]$ is a node in $p=(u,w)$, such that the distance between $u$ and $p[x]$ is $x$. The start and end nodes of the path are $p[0]=u$ and $p[e]=w$. We can calculate $p[x]$ in constant time as $p[x]=\LA(d(u)+x,w)$, where $d(u)$ is the depth of $u$, and $\LA$ is a Level Ancestor query.
\item The subpath $p[x\twodots y]=(p[x],p[y])$ of a path $p$ is the path, which starts at node $p[x]$ and ends at node $p[y]$. The entire path is $p = p[0,e]$.
\item The label of a path $\zlabel(p)=\zlabel(p[1])\cdot...\cdot\zlabel(p[e])$ is the concatenation of the labels on all edges between nodes in the path.
\end{itemize}

\paragraph{LCE on trees}
Given two paths $p_i$ and $p_j$ of lengths $e_i$ and $e_j$ in a tree $T$, the Longest Common Extension $v=\LCE_T(p_i, p_j)$ is the length of the longest common prefix of $\zlabel(p_i)$ and $\zlabel(p_j)$.

@@

@@

\fixme{Write out the query in full}

@@

\fixme{Improve this paragraph}
At any given iteration at level $\ell$, we know that $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and we now check if $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$. From our definition of fingerprints in a tree, these two labels have fingerprints $H_\ell[p_i[v+t_\ell]]$ and $H_\ell[p_j[v+t_\ell]]$, which is what our LCE query uses.
\end{proof}

@@

In a \proc{HeavyPathLce} query $\LCE_T(p_i,p_j)$ we start with $v=0$ and do the following repeatedly:
\begin{enumerate}
\item Stop when we reach either $v=e_i$ or $v=e_j$, and return $v$.
\item Find $m_i = \zlength(p_i[v], \NCA_T(\zend(p_i[v+1]), p_i[e_i]))$ \\ and $i=\zindex(p_i[v+1])$.
\item Find $m_j = \zlength(p_j[v], \NCA_T(\zend(p_j[v+1]), p_j[e_j]))$ \\ and $j=\zindex(p_j[v+1])$.
\item Find $v'=\min\{\LCE_s(i,j),m_i,m_j\}$.
\item Increment $v$ by $v'$.
\item Stop if $v'<\min\{m_i,m_j\}$ and return $v$.
\end{enumerate}

@@

The \proc{LadderLce} query starts with $v=0$ and repeats the following:
\begin{enumerate}
\item Stop when $v = e_i$ or $v = e_j$ and return $v$.
\item Find $m_i = \zlength(p_i[v\twodots e_i])/2$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
\fixme{afrunding}
\item Find $m_j = \zlength(p_j[v\twodots e_j])/2$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.
\item Find $v' = \min\{LCE_s(i,j),m_i,m_j\}$
\item Increment $v$ by $v'$.
\item Stop if $v' < \min\{m_i,m_j\}$ and return $v$.
\end{enumerate}
